---
title: "Bellabeat"
author: "Sabrina Vazquez"
date: "2023-02-19"
output: 
  html_document:
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: yes
    toc_float: yes
---

```{r}
library(tidyverse)
library(lubridate) 
library(ggplot2) 
library(janitor)
```

The function getwd() was employed to verify that the correct working directory was in use
```{r}
getwd()
```


Data was loaded
```{r}
Daily_Activity <- read_csv("Daily_Activity.csv")
Heart_Rate_Seconds <- read_csv("Heartrate_seconds.csv")
Hourly_Calories <- read.csv("Hourly_Calories.csv")
Hourly_Steps <- read.csv("Hourly_Steps.csv")
Hourly_Intensities <- read.csv("Hourly_Intensities.csv")
Sleep_Day <- read.csv("Sleep_Day.csv")
```

The "colnames()" function was utilized to obtain a brief overview of the data by examining the column names
```{r}
colnames(Daily_Activity)
colnames(Heart_Rate_Seconds)
colnames(Hourly_Calories)
colnames(Hourly_Steps)
colnames(Hourly_Intensities)
colnames(Sleep_Day)
```

How many distinct Id were in Heart_Rate_Seconds data set
```{r}
n_distinct(Heart_Rate_Seconds$Id)
```
#Only 14 users. Data was not used 

How many distinct Id were in Daily_Activity data set
```{r}
n_distinct(Daily_Activity$Id)
```

How many distinct Id were in Sleep_Day
```{r}
n_distinct(Sleep_Day$Id)
```
#less than 33 (24)

The following functions were used to collect specific insights about the Daily_Activity and Sleep_Day data:
```{r}
Daily_Activity %>%  
  select(TotalSteps,
         TotalDistance,
         SedentaryMinutes,
         VeryActiveMinutes) %>%
  summary()
```

```{r}
head(Daily_Activity)
head(Sleep_Day)
```

```{r}
str(Daily_Activity)
str(Sleep_Day)
```

ActivityDate column from Daily_Activity was changed into date format 
```{r}
Daily_Activity <- Daily_Activity %>% 
  mutate(ActivityDate= as_date(ActivityDate, format= "%m/%d/%Y"))
```

SleepDay column from Sleep_Day was changed to ActivityDate and changed into Date format
```{r}
Sleep_Day <- Sleep_Day %>%
  rename(ActivityDate= SleepDay) %>%
  mutate(ActivityDate= as_date(ActivityDate, format= "%m%d%y %I:%M%:S %p"))
```

Daily_Activity and Sleep_Day were merged
```{r}
Daily_Activity_Sleep <- merge(Daily_Activity, Sleep_Day, by= c("Id", "ActivityDate"))
```

How many distinct Id were in new Daily_Activity_Sleep data
```{r}
n_distinct(Daily_Activity_Sleep$Id) 
```
#24

Summary of the Data
```{r}
Daily_Activity_Sleep %>%  
  summary()
```

Visualization of Daily_Activity_Sleep
```{r}
ggplot(Daily_Activity_Sleep, aes(x=TotalSteps, y= Calories)) + 
  geom_point() +
  geom_jitter() +
  geom_smooth(color = "#800080") +
  labs(title= "Total Steps vs Calories Burnt")
```

```{r}
ggplot(Daily_Activity_Sleep, aes(x=VeryActiveMinutes, y= Calories)) + 
  geom_point() +
  geom_jitter() +
  geom_smooth(color = "#800080") +
  labs(title= "Very Active Minutes vs Calories Burnt")
```

```{r}
ggplot(Daily_Activity_Sleep, aes(x=TotalMinutesAsleep, y=SedentaryMinutes)) + 
  geom_point() +
  geom_jitter() +
  geom_smooth(color = "#800080") +
  labs(title= "Total Minutes Asleep vs Sedentary Minutes")
```

```{r}
ggplot(data=Daily_Activity_Sleep, aes(x=TotalMinutesAsleep, y=TotalTimeInBed)) + 
  geom_point() + 
  geom_smooth(color = "#800080") +
  labs(title="Total Minutes Asleep vs Total Time in Bed")
```

Columns that list the date, month, day, and year were added
```{r}
Daily_Activity_Sleep <- Daily_Activity_Sleep %>% 
  mutate(month = format(as.Date(ActivityDate), "%B"),
         year = year(ActivityDate),
         day = day(ActivityDate),
         day_of_week = wday(ActivityDate, label = TRUE))
```

A new data set was created to get the avg. daily steps and avg. daily sleep per weekday
```{r}
Steps_Sleep_By_Weekday <- Daily_Activity_Sleep%>%
  group_by(day_of_week) %>%
  summarize (daily_steps = mean(TotalSteps), daily_sleep = mean(TotalMinutesAsleep))
```

Visualization of Steps_Sleep_By_Weekday
*Steps
```{r}
Steps_Sleep_By_Weekday %>% 
  ggplot(aes(x = day_of_week, y = daily_steps)) +
  geom_col(fill = "#800080", position = "dodge") +
  labs(title="Avg. Daily Steps by Weekday")
```

*Sleep
```{r}
Steps_Sleep_By_Weekday %>% 
  ggplot(aes(x = day_of_week, y = daily_sleep)) +
  geom_col(fill = "#9370DB", position = "dodge") +
  labs(title="Avg. Daily Sleep by Weekday")
```

A new data set was created to get the daily averages
```{r}
Daily_Averages <- Daily_Activity_Sleep %>%
  group_by(Id) %>%
  summarise (avg_daily_steps = mean(TotalSteps), avg_daily_calories = mean(Calories), avg_daily_sleep = mean(TotalMinutesAsleep))
```

```{r}
Daily_Averages %>%  
  summary()

head(Daily_Averages)
```

A new column was added to the dataset, which classifies users into different categories based on the amount of daily steps taken. 
```{r}
Daily_Averages <- Daily_Averages %>% 
  mutate(user_type = case_when(
    avg_daily_steps < 5000 ~ "Sedentary",
    avg_daily_steps >= 5000 & avg_daily_steps < 9999 ~ "Lightly Active", 
    avg_daily_steps >= 10000 & avg_daily_steps < 14999 ~ "Fairly Active", 
    avg_daily_steps >= 15000 ~ "Very Active"
  ))
```

Summary of the frequency of each unique value of the user_type column in the Daily_Averages data frame.
```{r}
user_type_frequencies <- Daily_Averages %>%
  count(user_type) %>%
  mutate(percent = round(n / sum(n) * 100, 1),
         percent_label = paste0(percent, "%"))
```

Visualization
```{r}
user_type_frequencies %>%
  ggplot(aes(x="",y=percent, fill=user_type)) +
  geom_bar(stat = "identity", width = 1)+
  coord_polar("y", start=0)+
  geom_text(aes(label = percent_label, fontface="bold"),
            position = position_stack(vjust = 0.5))+
  labs(title="User Type distribution") +
  theme_minimal()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size=14, face = "bold")) +
  scale_fill_manual(values = c("#800080", "purple4", "#9370DB","#9400D3"))
```

A new dataset was generated, consisting of two columns. The first column denotes the days in which the device was utilized, while the second column indicates the level of usage.
```{r}
Daily_Use <- Daily_Activity_Sleep %>%
  group_by(Id) %>%
  summarize(days_used=sum(n())) %>%
  mutate(usage = case_when(
    days_used >= 1 & days_used <= 10 ~ "Low",
    days_used >= 11 & days_used <= 20 ~ "Moderate", 
    days_used >= 21 & days_used <= 32 ~ "High", 
  ))
```

Summary of the frequency of each unique value of the usage column in the Daily_Use data frame
```{r}
usage_frequencies <- Daily_Use %>%
  count(usage) %>%
  mutate(percent = round(n / sum(n) * 100, 1),
         percent_label = paste0(percent, "%"))
```

Visualization
```{r}
usage_frequencies %>%
  ggplot(aes(x="",y=percent, fill=usage)) +
  geom_bar(stat = "identity", width = 1)+
  coord_polar("y", start=0)+
  geom_text(aes(label = percent_label, fontface="bold"),
            position = position_stack(vjust = 0.5))+
  labs(title="Usage frequency") +
  theme_minimal()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size=14, face = "bold")) +
  scale_fill_manual(values = c("#800080", "purple4", "#9370DB"))
```




